Combinatorial interpretations of holonomic recurrences

### 2021-09-08 article/main.tex
Add more stories

\section{Overview}%
\todo{generating functions}
\todo{holonomic recurrences}
\todo{publications about Motzkin, etc.}
\todo{Lambda terms}
\todo{zigzag free}

Every recursive structure can be described using a~generating function, which can be then transformed into a~holonomic recurrence. It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding vertices. However, there is no known method of interpreting it in the language of the combinatorial structures.

The initial inspiration for the thesis came from the publication \textit{A natural counting of lambda terms}~\cite{inspiration}, which shows bijections among seemingly different tree structures, which are lambda trees, black-white trees~\cite{blackwhite} and zigzag free trees\todo{cite}. They share the same holonomic recurrence, but there is no known immediate connection between the holonomic recurrence and the structures itself.

@@

Next, in Section~\ref{sec:known_results}, we present known results, ie. binary trees, unary-binary trees and Schröder trees. They demonstrate combinatorial interpretations that reflect different approaches.

@@

\begin{figure}[H]
    \begin{center}
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B\)\\
                \includesvg[scale=0.5]{binary__def_1}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(=\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(\text{leaf}\)\\
                \includesvg[scale=0.5]{binary__def_2}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(+\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B^2\)\\
                \includesvg[scale=0.5]{binary__def_3}%
            \end{center}
        \end{minipage}%
    \end{center}
    \caption{Binary tree recursive definition}
    \label{fig:binary_recursion}
\end{figure}

@@

\begin{definition}
The ordinary generating function of sequence \(A = (a_i)_{i=0}^{\infty}\) is defined as a~formal power series:
\[\gf{A}(z) = \sum_{i=0}^{\infty} a_i z^i .\]
\end{definition}

@@

\begin{example}
    \label{ex:bin_gf}
    Let's consider recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can create a~functional equation to count trees with exactly n internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

@@

\[\begin{array}{rclr}
        \gf{B}(z) &=& 1 + z\gf{B}^2(z)\\
                  &=& 1 + z\bigg(1 + z\gf{B}^2(z)\bigg)\bigg(1 + z\gf{B}^2(z)\bigg)\\
                  &=& 1 + \mhl{z}\bigg(1 + \mhl{z}\Big(\mhl{1} + z\gf{B}^2(z)\Big)\Big(\mhl{1} + z\gf{B}^2(z)\Big)\bigg)\bigg(\mhl{1} + z\gf{B}^2(z)\bigg) &(*)\\
                  &=& 1 + z\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\\
                  &=& 1 + z + 2 z^2 + \bigg(2 \gf{B}^2(z) + \Big(1 + z \gf{B}^2(z)\Big)^2\bigg) z^3
\end{array}\]

>>>>

\section{Overview}%

Generating functions is a way of describing infinite sequences using formal power series, where coefficients are encoding the the consecutive elements. It allows us to incorporate many advanced methods of analysis of encoded objects.

Every recursive structure can be described using a~generating function equation, which can be then solved and transformed into a~holonomic recurrence, i.e.~recurrence that in order to define \(n\)-th element of the sequence requires just finite number of previous ones and coefficients that are polynomials of \(n\).

It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding vertices. However, there is no known method of interpreting it in the language of the combinatorial structures.

Up to today only a few interpretations of holonomic recurrences on trees have been found, including interpretation of holonomic recurrences for binary trees~\cite{binary}, unary-binary trees~\cite{motzkin}, and Schröder trees~\cite{schroder}. Unfortunately, the mentioned interpretations do not come with a method of finding them that could be generatlized to more complex cases.

The initial inspiration for the thesis came from the publication \textit{A natural counting of lambda terms}~\cite{inspiration}, which shows bijections among seemingly different tree structures, which are lambda trees, black-white trees~\cite{blackwhite} and zigzag free trees~\cite{inspiration}\todo{Find better cite}. They share the same holonomic recurrence, but there is no known immediate connection between the holonomic recurrence and the structures itself.

Lambda terms are very popular structures used for analyzing functions and programming languages. Their very simple recursive definition makes them very attractive. Better understanding them could allow us improving programming language compilers and in consequence improve every aspect of our lives. The interpretation of holonomic recurrence for lambda trees itself could allow us to discover new ways of generating random lambda terms via extending smaller ones in similar way to Rémy's algorithm for sampling binary trees~\cite{remy,note}.

@@

Next, in Section~\ref{sec:known_results}, we present known results, i.e.~binary trees, unary-binary trees and Schröder trees. They demonstrate combinatorial interpretations that reflect different approaches.

@@

\begin{figure}[H]
    \begin{center}
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B\)\\
                \includesvg[scale=0.5]{binary__def_1}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(=\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(\text{leaf}\)\\
                \includesvg[scale=0.5]{binary__def_2}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(+\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B^2\)\\
                \includesvg[scale=0.5]{binary__def_3}%
            \end{center}
        \end{minipage}%
    \end{center}
    \caption{Binary tree recursive definition}%
    \label{fig:binary_recursion}
\end{figure}

@@

\begin{definition}
    The ordinary generating function of sequence \(A = {(a_i)}_{i=0}^{\infty}\) is defined as a~formal power series:
\[\gf{A}(z) = \sum_{i=0}^{\infty} a_i z^i.\]
\end{definition}

@@

\begin{example}%
    \label{ex:bin_gf}
    Let's consider recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can create a~functional equation to count trees with exactly n internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

@@

\[\begin{array}{rclr}
        \gf{B}(z) &=& 1 + z\gf{B}^2(z)\\
                  &=& 1 + z\bigg(1 + z\gf{B}^2(z)\bigg)\bigg(1 + z\gf{B}^2(z)\bigg)\\
                  &=& 1 + \mhl{z}\bigg(1 + \mhl{z}\Big(\mhl{1} + z\gf{B}^2(z)\Big)\Big(\mhl{1} + z\gf{B}^2(z)\Big)\bigg)\bigg(\mhl{1} + z\gf{B}^2(z)\bigg) &(*)\\
                  &=& 1 + z\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\\
                  &=& 1 + z + 2 z^2 + \bigg(2 \gf{B}^2(z) + {\Big(1 + z \gf{B}^2(z)\Big)}^2\bigg) z^3
\end{array}\]
